---
title: "Parameter estimation and sampling methods"
date: "September 12, 2018"
output: 
    html_document:
        self_contained: false
        lib_dir: libs
        theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

In this lab, you will apply the concepts seen during the last two classes.


## 1. Flower measurements of three iris species

For this exercise, we will use Edgar Anderson's famous iris data set, which contains different measurements taken on 50 flowers of 3 iris species. This data frame is already loaded into R under the name `iris`.

```{r}
head(iris)
```

a) We will first visualize some of the data. Using *ggplot2*, produce a scatter plot relating the length and width of the sepals (`Sepal.Length` and` Sepal.Width`), differentiating the points of each species by color.

How would you calculate the mean of `Sepal.Width` and its 95% confidence interval by species?

b) What quantities do you need for this calculation?

c) Using the *dplyr* package, calculate the mean, sample size, standard deviation and standard error of the mean of `Sepal.Width` for each species (`Species`). Save the result in a data frame `iris_stat`.

d) During the class on statistical distributions, we saw the functions `rnorm`,` dnorm`, `pnorm` and` qnorm` which calculate values from the normal distribution. Similar functions exist for the $t$ distribution (`rt`,` dt`, `pt`,` qt`). Let's use the function `qt(p, df)` to determine the interval corresponding to 95% of the probability. What values of $p$ (cumulative probability) should we use? How many degrees of freedom ($df$) based on sample size $n$?

e) Create two new columns in `iris_stat` containing the minimum` ic_min` and the maximum `ic_max` of the confidence interval. Define these columns as functions of the mean, standard error and sample size.

f) Finally, use the `geom_pointrange` graph type of *ggplot2* to visualize the confidence interval for each species. This type of chart requires the specification of `y` (center point),` ymin` (minimum of range) and `ymax` (maximum of range) in the` aes` function.


## 2. Mean DBH of species sampled at Kejimkujik National Park

Based on the code written in the last exercise, produce a graph of the mean DBH with the 95% confidence interval for all species in the data frame from the file `cours1_kejimkujik.csv`. Then, from the data and graph, answer the following questions.

a) What is the confidence interval of the mean DBH for white pine (*PIST*)? Assuming that the sampled individuals are representative of the population on this site, how do you interpret this interval?

b) If the sample is representative of each species, can it be easily determined which species has the smallest mean DBH, or which has the highest mean DBH, in this population?


## 3. *t* distribution

We will visually compare the properties of the *t* distribution and the standard normal distribution.

a) Create a `tab1` data frame with a single column,`x` that contains all values between -3 and 3, at 0.1 intervals.

b) Add a `z` column that contains the density value of a standard normal distribution for each value of` x`. If you are unsure of the R function to get the density of a normal variable, use the `?Normal` help command.

c) Add a second column `t9` that contains the density value for the *t* distribution with 9 degrees of freedom.

d) Create a graph of `z` vs. `x` with` geom_line`, and assign it to an object in R.

e) Add to the saved graph a different color curve for `t9`.
You can add a line where `y` is associated with a different variable as follows:
`[saved chart name] + geom_line(aes(y = t9), color = "blue")`.

f) Use `rt` to produce a sample of 100 values of the *t* distribution at 9 degrees of freedom, then compare this vector to a normal distribution with a quantile-quantile graph.

g) If we took an interval containing 80% of the probability for each of the two distributions (`z` and` t9`), which interval would be the widest? Try to deduce the answer from the graphs obtained in e) and f), then check by calculating the quantiles corresponding to the 80% interval.


## 4. Choice of sampling methods

Imagine you want to determine the mean and standard deviation of the growth rate of yellow birch seedlings in a given region. Your sampling unit is a 1 m$^2$ quadrat where the annual growth of seedlings is measured. Which sampling method would you recommend for the placement of these quadrats according to the different scenarios presented, and why? You can choose stratified, cluster, systematic, or adaptive sampling.

a) The stands of this species in the area are relatively similar but far apart.

b) Growth could be strongly influenced by the north-south temperature gradient at the scale considered.

c) The species is found in several distinct stand types covering different proportions of the territory (eg 70% type A, 25% type B, 5% type C).

d) It is a rare species in the region and its distribution is little known.


## 5. Stratified sampling simulation

For this exercise, we will compare simple and stratified sampling from simulated samples from the iris data frame. Here is the distribution of the 50 petal lengths for each species in the original data frame.

```{r}
ggplot(iris, aes(x = Species, y = Petal.Length)) + 
    geom_boxplot()
```

In *dplyr*, the `sample_n(df, n)` function returns a data frame containing `n` randomly selected observations from the data frame `df`. It can also be used with `group_by` to choose` n` observations by group.

a) Create two functions `iris_alea` and` iris_strat`. The first function chooses 30 random observations of `iris`, then returns the mean of `Petal.Length` for these observations. The second chooses 10 random observations from each of the three species, then returns the mean of `Petal.Length` (overall mean, not by species). Make sure that each of the two functions returns a vector of length 1.

*Note*: You can write these functions without arguments (empty parentheses after `function`), as in the example below.

```{r eval = FALSE}
iris_alea <- function() {
    # Insérer code de la fonction ici
}

```

b) Generate a vector of 1000 results of each function with `replicate`, as follows:

```{r eval = FALSE}
rep_alea <- replicate(1000, iris_alea())
rep_strat <- replicate(1000, iris_strat())
```

c) Calculate the standard error of each mean (from the standard deviation of `rep_alea` and `rep_strat`). Before running the calculation, can you predict which method will be more precise and why?


