---
title: "Hypothesis tests on the mean"
date: "September 19, 2018"
output: 
    html_document:
        self_contained: false
        lib_dir: libs
        theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
```

## 1. Hypothesis tests

Which of the following statements are true?

a) The $p$ value is the probability that the null hypothesis is true.

(b) The higher the significance threshold $\alpha$, the more difficult it is to reject the null hypothesis.

c) For a one-sample $t$-test, if we reject the null hypothesis $\mu = 10$ with a threshold $\alpha = 0.01$, then the 99% confidence interval of the sample mean $\bar{x}$ does not contain the value 10.

d) For the $t$-test comparing two samples with the null hypothesis $\mu_A = \mu_B$, a smaller value $p$ means that there is a greater difference between the means.

e) In experimental design, random assignment of treatments is intended to ensure that the differences measured between groups are due to the treatment.


## 2. Ozone concentration in three gardens

For this exercise, we will use the `gardens.csv` dataset, which comes from Michael Crawley's book *Statistics: An Introduction Using R*. These data represent ozone concentrations (in parts per 100 million or pphm) measured in three gardens (A, B and C) on different days.

```{r}
gardens <- read.csv("gardens.csv")
```

a) Observe the distribution of ozone measurements in each garden. What type of graph could you use?

b) What is the mean and standard deviation of the ozone concentration in each garden? Is the mean a good indicator of the "typical" value in each garden?

c) From these data, test the null hypothesis that gardens A and B receive the same mean ozone concentration. What is your estimate of the difference between the mean and its 99% confidence interval? Does this test give a good idea of the difference between the two gardens? Explain your answer.

d) Repeat the previous exercise for the null hypothesis that gardens A and C receive the same average ozone concentration. Comment on the difference between this result and the previous result.


## 3. Display the estimate of a *t*-test

We can save the result of a $t$-test in a variable, for example:

```{r}
res_t <- t.test(Ozone ~ Garden, data = filter(gardens, Garden != "C"),
       conf.level = 0.99)
```

Then, we can access different parts of that result with the `$` operator.

```{r}
res_t$p.value
res_t$estimate
```

Create a function that takes one argument, which is the result of a $t$-test (performed on 2 samples), and that displays the estimate of the difference between the means with its confidence interval. Round the numbers to one decimal place. For example, for the result `res_t` above, the function should produce the text: "-2 (-3.5, -0.5)".

**Hint**

- Use the `round` function to round a numeric variable.

- The `paste0` function allows you to paste numbers and character strings together. For example:

```{r}
t <- 6
paste0("It is ", t, " o'clock.")
```

*Note*: The function `paste` works in the same way, but adds a space between each element.


## 4. Experiment with paired samples

Give one example of an experiment where the results could be analyzed with a paired sample $t$-test.